---
title: "Clones"
output:
  html_notebook:
    toc: yes
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    toc_float: true
    theme: united  # many options for theme, this one is my favorites.
    highlight: tango  # specifies the syntax highlighting style
    code_folding: hide
#editor_options: 
  #chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
HOME <- "~/Library/CloudStorage/Dropbox/Steffi_Zostera_analysis_2023/Monitoring2023/Monitoring_eelgrass_scripts_for_figshare" # directory 
DATA <- paste0(HOME, "/data/") # data folder
RESULT <- paste0(HOME, "/results/") #result folder
SCRATCH <- paste0(HOME, "/scratch/") #result folder
```

```{r include=FALSE}
library(poppr)
library(vcfR)
library(adegenet)
library(poppr)
library(ape)
library(RColorBrewer)
library(fishualize)
library(igraph)
library(phangorn)
```

# Load data
```{r}
#load genind object
load(paste0(SCRATCH, "gl_zostera_25.rda"))
load(paste0(SCRATCH, "gen_zostera_25.rda"))

gl_zostera <- gl_zostera_25 # rename to work with following scripts
gen_zostera <- gen_zostera_25 # rename to work with following scripts
```


```{r, fig.height=25, fig.width=10}
reg_col <- fish(n = nlevels(gl_zostera$other$ind.metrics$region), option = "Hypsypops_rubicundus")[gl_zostera$other$ind.metrics$region]
pop_col <- fish(n = nPop(gl_zostera), option = "Hypsypops_rubicundus")[gl_zostera$pop]
```

# MLGs
Calculate distance
```{r}
#calculate genetic distance 
bit_dist <- bitwise.dist(gl_zostera) # Hamming distance, or the number of differences between two strings.
```

```{r}
#convert to snpclone
sc<-as.snpclone(gl_zostera)
# Naive (original) MLG
nmll(sc)
```

# Hammings distance - bitwise.dist

```{r,fig.height=5, fig.width=10}
# for some reason plotting becomes weird with bitwise.dist
eelgrass_filtered <- filter_stats(sc, distance = bitwise.dist,  plot = TRUE, hist = "scott")
```

```{r,fig.height=5, fig.width=10}
# for some reason plotting becomes weird with bitwise.dist
# plotting the genetic distance cutoff with ablines
# each column is 0.001 of Genetic Distance Cutoff
#png(paste0(RESULT, "bitwise.dist_genetic.distance_cutoff.png"), width = 20, height = 10, units = "cm", res = 300) 

eelgrass_filtered <- filter_stats(sc, distance = bitwise.dist,  plot = TRUE, hist = "scott")
  cex <- 1.5
  par(cex.lab=cex, cex.axis=cex, cex.main=cex)
  abline(v = 0.01543498, col = "black", lty=1)
  legend("bottomright", 
       legend = c("Max replicate distance"),
       pch = "|",
       col = c("black"),
       title = "Thresholds")
#dev.off()
```

# Different thresholds
```{r}
#this if for later plotting the tree
max_thresh <- max(eelgrass_filtered$average$THRESHOLDS)/2
max_thresh

```

## distance based of replicates
```{r}
library(tidyverse)

# takes distance matrix and makes it into a tibble anf filter to only have replicate pairs
df_dist <- as.matrix(bit_dist) %>% 
  as_tibble(rownames = "ind1") %>% 
  pivot_longer(!ind1, names_to = "ind2", values_to = "dist") 
rep_dist <- df_dist %>% 
  filter(str_detect(ind1, "rep$")) %>% 
  filter(!str_detect(ind2, "rep$")) %>% 
  mutate(ind1match = str_replace(ind1, "-rep","")) %>% 
  filter(ind2 == ind1match) %>% 
  dplyr::select(ind1 = ind1, ind2 = ind2, dist = dist) %>%
  arrange(desc(dist))

#(max_rep_dist <- max(rep_dist$dist))
max_rep_dist <- 0.01543498
```

```{r, fig.height=10, fig.width=10}
# filter now with the max rep dist: 0.01543499
max_rep_dist <- 0.01543498
mlg.filter(sc, distance = bitwise.dist) <- max_rep_dist
nmll(sc)
sc_mlg <- mlg.table(sc)
```

```{r}
#bitwise_dist_matrix <- bitwise.dist(gl_zostera, mat = T)
#head(bitwise_dist_matrix)
#write.csv(bitwise_dist_matrix, "bitwise_dist_matrix.csv", row.names = TRUE)
```

## Tree

```{r, cache=TRUE}
#distance tree
set.seed(9)
bit_tree_2 <- aboot(gl_zostera, tree = "wpgma", distance = bitwise.dist, sample = 100, showtree = F, cutoff = 50, quiet = T)
save(bit_tree_2, file=paste0(SCRATCH,"bit_tree_2.rda"))
```

```{r}
load(file = paste0(SCRATCH,"bit_tree_2.rda"))
```

# THE TREE TO PRINT - WPGMA TREE - SHOWS US MORE RELIABLE CLADES
```{r, fig.height=50, fig.width=10}
#png(paste0(RESULT, "bit_wpgma_tree.png"), width = 10, height = 50, units = "in", res = 300)
plot.phylo(bit_tree_2, 
           cex = 1, 
           main = 'Genetic distance',
           font = 2, 
           adj = 0, 
           tip.color =  pop_col)
axisPhylo()
nodelabels(bit_tree_2$node.label, adj = c(1.3, -0.5), frame = "n", cex = 0.8,font = 3, xpd = TRUE)
legend("topleft", 
       legend = unique(gl_zostera$other$ind.metrics$site_full),
       fill = fish(n = 15, option = "Hypsypops_rubicundus"),
       title = "Sites")
legend("topright", 
       legend = c("Max replicate distance"),
       pch = "|",
       col = c("black"),
       title = "Thresholds")
abline(v = max_thresh-max_rep_dist/10, col = c("black"), lty=2)
#dev.off()
```


# Stats without replicates
```{r, include=FALSE}
eelgrass_filtered <- filter_stats(sc, distance = bitwise.dist,  plot = TRUE, hist = "scott")
#average_thresh  <- cutoff_predictor(eelgrass_filtered$average$THRESHOLDS)
mlg.filter(sc, distance = bitwise.dist) <- max_rep_dist
nmll(sc)
```

```{r}
# removing replicates
sc_no_rep <- sc[!(indNames(sc) %in% rep_dist$ind1),]
save(sc_no_rep, file=paste0(SCRATCH, "sc_no_rep.rda"))
```

```{r}
sc_no_rep_mlg <- clonecorrect(sc_no_rep, strata = ~Site)
popNames(sc_no_rep_mlg) <- popNames(sc_no_rep)
save(sc_no_rep_mlg, file=paste0(SCRATCH, "sc_no_rep_mlg.rda"))
```

```{r, fig.height=10, fig.width=10}
sc_mlg <- mlg.table(sc_no_rep)
```


```{r}
#Function for number of individuals
nInd <- function(x){
 x <- drop(as.matrix(x))
 if (length(dim(x)) > 1){ # if it's a matrix
   res <- rowSums(x)
 } else {                 # if it's a vector
   res <- sum(x)
 }
 return(res)
}

#Function for number of MLGs
nMLG <- function(x){
 x <- drop(as.matrix(x))
 if (length(dim(x)) > 1){ # if it's a matrix
   res <- rowSums(x>0)
 } else {                 # if it's a vector
   res <- sum(x>0)
 }
 return(res)
}

#Function for calculating clonal richness
nCR <- function(x){
 x <- drop(as.matrix(x))
 if (length(dim(x)) > 1){ # if it's a matrix
   res <- (rowSums(x > 0)-1)/(rowSums(x)-1)
 } else {                 # if it's a vector
   res <- (sum(x > 0)-1)/(sum(x)-1)
 }
 return(res)
}
```

```{r}
div_stat <- diversity_stats(sc_mlg, nInd = nInd, nMLG = nMLG, nCR = nCR)
div_stat
```

```{r}
#remove replicates from genlight
gl_no_rep <- gl_zostera[!(indNames(gl_zostera) %in% rep_dist$ind1),]

library(dartR)
het_df <- gl.report.heterozygosity(gl_no_rep)
```


## Stats with poppr on genclone N
Here using geneclone instead as it works better for calculating statistics, but should note that the distances change, but not the number of MLGs..

```{r, cache = TRUE}
gc <- as.genclone(gen_zostera)
gc_custom_thresh <- 65
mlg.filter(gc, distance = diss.dist) <- gc_custom_thresh
gc_mlg1 <- mlg.table(gc)
gc[!(indNames(gc) %in% rep_dist$ind1),]
```


```{r, cache=TRUE , fig.height=4, fig.width=5}
# remove replicates
gc_no_rep <- gc[!(indNames(gc) %in% rep_dist$ind1),]
save(gc_no_rep, file=paste0(SCRATCH, "gc_no_rep.rda"))

#to check mll
gc_mlg <- mlg.table(gc_no_rep)
```

```{r}
gc_no_rep_mlg <- clonecorrect(gc_no_rep, strata = ~Site)
popNames(gc_no_rep_mlg) <- popNames(gc_no_rep)
save(gc_no_rep_mlg, file=paste0(SCRATCH, "gc_no_rep_mlg.rda"))
```

```{r, include=FALSE, cache=TRUE}
# diversity stats
gc_stats <- poppr(gc_no_rep)
save(gc_stats, file = paste0(SCRATCH,"gc_diversity_stats.rda"))
```

```{r}
library(kableExtra)
load(file = paste0(SCRATCH,"gc_diversity_stats.rda"))

basic_stats<- gc_stats %>% 
  as_tibble() %>% 
  mutate(nCR = (MLG-1)/(N-1))

basic_stats %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

write.csv(basic_stats, file = paste0(RESULT, "basicstats.csv"))
```

```{r, fig.height=8, fig.width=10}
basic_stats$Pop <- factor(basic_stats$Pop, levels= unique(basic_stats$Pop))
basic_stats %>% 
  as.tibble() %>% 
  pivot_longer(-c(Pop, File)) %>% 
  filter(Pop != "Total") %>% 
  ggplot(aes(x = Pop, y = value, col = Pop)) +
    geom_point() +
    facet_wrap(facets = vars(name), scales = "free") +
  expand_limits(x = 0, y = 0) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
```

## Stats with poppr on genclone MLG

```{r, cache=TRUE}
# diversity stats
gc_stats_MLG <- poppr(gc_no_rep, clonecorrect = TRUE, strata = ~Site)
save(gc_stats_MLG, file = paste0(SCRATCH,"gc_diversity_stats_MLG.rda"))
```

```{r}
library(kableExtra)
load(file = paste0(SCRATCH,"gc_diversity_stats_MLG.rda"))

basic_stats_MLG<- gc_stats_MLG %>% 
  as.tibble() %>% 
  mutate(nCR = (MLG-1)/(N-1))

basic_stats_MLG %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

write.csv(basic_stats_MLG, file = paste0(RESULT, "basicstats_MLG.csv"))
```

```{r, fig.height=8, fig.width=10}
basic_stats_MLG$Pop <- factor(basic_stats_MLG$Pop, levels= unique(basic_stats_MLG$Pop))
basic_stats_MLG %>% 
  as.tibble() %>% 
  pivot_longer(-c(Pop, File)) %>% 
  filter(Pop != "Total") %>% 
  ggplot(aes(x = Pop, y = value, col = Pop)) +
    geom_point() +
    facet_wrap(facets = vars(name), scales = "free") +
  expand_limits(x = 0, y = 0) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
```

# LINKAGE DISEQUILIBRIUM

This test is useful to determine if populations are clonal (where significant disequilibrium is expected due to linkage among loci) or sexual (where linkage among loci is not expected)
```{r}
library("poppr")
library("magrittr")

# load data -> fir ia() we need genind or genclone object -> genclone didnt work
load(paste0(SCRATCH, "gc_no_rep.rda"))
load(paste0(SCRATCH, "gc_no_rep_mlg.rda"))

LD_gd <- genclone2genind(gc_no_rep)

LD_gd_MLG <- genclone2genind(gc_no_rep_mlg)
```

## calculate LD for each coast - compare coasts with and without clones

### west coast
```{r}
set.seed(1)
LD_west <- popsub(LD_gd, sublist = "ALL", exclude = c("YST", "FUR", "KAR", "HOR", "KLI", "SLI", "KRA", "NYN", "BJO"))
LD_plot_west <- ia(LD_west, sample = 999)

# we see stats close to 0: r = 0.0124 which doesnt fall into the linkage disequilibrium and shows us that these samples reproduce sexually
```

```{r}
set.seed(2)
LD_west_MLG <- popsub(LD_gd_MLG, sublist = "ALL", exclude = c("YST", "FUR", "KAR", "HOR", "KLI", "SLI", "KRA", "NYN", "BJO"))
LD_plot_west_MLG <-ia(LD_west_MLG, sample = 999)
```

### east coast
```{r}
# takes a very long time... couldn't figure out how to speed up
set.seed(1)
LD_east <- popsub(LD_gd, sublist = "ALL", exclude = c("KOD", "STE", "GOT", "GRO", "HOG", "ALA"))
LD_plot_east <- ia(LD_east, sample = 999)
```

```{r}
set.seed(2)
LD_east_MLG <- popsub(LD_gd_MLG, sublist = "ALL", exclude = c("KOD", "STE", "GOT", "GRO", "HOG", "ALA"))
LD_plot_east_MLG <- ia(LD_east_MLG, sample = 999)
```

## calculate LD for each west site - compare sites with and without clones

### KOD
```{r}
# clones no reps
LD_KOD <- popsub(LD_gd, "KOD")
ia(LD_KOD, sample = 999)

#          Ia        p.Ia       rbarD        p.rD                                                                               
# 28.42187450  0.00100000  0.06968563  0.00100000 
```

```{r}
# MLG no reps
LD_KOD_MLG <- popsub(LD_gd_MLG, "KOD")
ia(LD_KOD_MLG, sample = 999)

#        Ia       p.Ia      rbarD       p.rD          
#6.40643932 0.02500000 0.01627867 0.02300000 
```

### STE
```{r}
# clones no reps
LD_STE <- popsub(LD_gd, "STE")
ia(LD_STE, sample = 999)

#          Ia        p.Ia       rbarD        p.rD                                                                               
# 5.468466438 0.001000000 0.009585033 0.001000000 
```

```{r}
LD_STE_MLG <- popsub(LD_gd_MLG, "STE")
ia(LD_STE_MLG, sample = 999)

#          Ia         p.Ia        rbarD         p.rD                                                                                                                                              
# 0.4920443227 0.0390000000 0.0008563595 0.0390000000 
```

### GOT
```{r}
# clones no reps
LD_GOT <- popsub(LD_gd, "GOT")
ia(LD_GOT, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                                                                                                                  
# 0.762711864 0.102000000 0.003598616 0.100000000 
```

```{r}
LD_GOT_MLG <- popsub(LD_gd_MLG, "GOT")
ia(LD_GOT_MLG, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                                                                                                            
# 0.762711864 0.125000000 0.003598616 0.118000000
```

### GRO
```{r}
# clones no reps
LD_GRO <- popsub(LD_gd, "GRO")
ia(LD_GRO, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                     
# 43.5636542  0.0010000  0.4150484  0.0010000
```

```{r}
LD_GRO_MLG <- popsub(LD_gd_MLG, "GRO")
ia(LD_GRO_MLG, sample = 999)

#    Ia  p.Ia rbarD  p.rD 
#   NA    NA    NA    NA 
```

### HOG
```{r}
# clones no reps
LD_HOG <- popsub(LD_gd, "HOG")
ia(LD_HOG, sample = 999)

#          Ia        p.Ia       rbarD        p.rD                                                            
# 12.36183150  0.00100000  0.03354375  0.00100000 
```

```{r}
LD_HOG_MLG <- popsub(LD_gd_MLG, "HOG")
ia(LD_HOG_MLG, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                            
# 1.256800046 0.003000000 0.003471862 0.003000000
```


### ALA
```{r}
# clones no reps
LD_ALA <- popsub(LD_gd, "ALA")
ia(LD_ALA, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                
# 6.59209857 0.00100000 0.01410926 0.00100000
```

```{r}
LD_ALA_MLG <- popsub(LD_gd_MLG, "ALA")
ia(LD_ALA_MLG, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                            
# 1.434140043 0.121000000 0.003364167 0.133000000 
```

## calculate LD for each east site - compare sites with and without clones

### YST
```{r}
# clones no reps
LD_YST <- popsub(LD_gd, "YST")
ia(LD_YST, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                                  
# 22.42565715  0.00100000  0.09808751  0.00100000 
```

```{r}
LD_YST_MLG <- popsub(LD_gd_MLG, "YST")
ia(LD_YST_MLG, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                                  
# 1.069237511 0.554000000 0.006120415 0.557000000 
```

### FUR
```{r}
# clones no reps
LD_FUR <- popsub(LD_gd, "FUR")
ia(LD_FUR, sample = 999)

#          Ia        p.Ia       rbarD        p.rD                                                            
# 105.3359538   0.0010000   0.7221526   0.0010000 
```

```{r}
LD_FUR_MLG <- popsub(LD_gd_MLG, "FUR")
ia(LD_FUR_MLG, sample = 999)

# Ia  p.Ia rbarD  p.rD 
# NA    NA    NA    NA 
```

### KAR
```{r}
# clones no reps
LD_KAR <- popsub(LD_gd, "KAR")
ia(LD_KAR, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                
# 34.1007156  0.0010000  0.1873399  0.0010000 
```

```{r}
LD_KAR_MLG <- popsub(LD_gd_MLG, "KAR")
ia(LD_KAR_MLG, sample = 999)

#   Ia  p.Ia rbarD  p.rD 
#   NA    NA    NA    NA 
```

### KLI
```{r}
# clones no reps
LD_KLI <- popsub(LD_gd, "KLI")
ia(LD_KLI, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                            
# 11.79841851  0.00100000  0.05061003  0.00100000 
```

```{r}
LD_KLI_MLG <- popsub(LD_gd_MLG, "KLI")
ia(LD_KLI_MLG, sample = 999)

#         Ia        p.Ia       rbarD        p.rD                                                            
# 1.457461407 0.018000000 0.006986258 0.018000000 
```

### HOR
```{r}
# clones no reps
LD_HOR <- popsub(LD_gd, "HOR")
ia(LD_HOR, sample = 999)

#        Ia       p.Ia      rbarD       p.rD                                                            
# 9.84669959 0.00100000 0.05335167 0.00100000
```

```{r}
LD_HOR_MLG <- popsub(LD_gd_MLG, "HOR")
ia(LD_HOR_MLG, sample = 999)

#    Ia  p.Ia rbarD  p.rD 
#   NA    NA    NA    NA 
```

### SLI
```{r}
# clones no reps
LD_SLI <- popsub(LD_gd, "SLI")
ia(LD_SLI, sample = 999)

#        Ia       p.Ia      rbarD       p.rD                                                                
# 30.2822954  0.0010000  0.1281995  0.0010000
```

```{r}
LD_SLI_MLG <- popsub(LD_gd_MLG, "SLI")
ia(LD_SLI_MLG, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                
# 38.3392857  0.0010000  0.2665094  0.0010000 
```

### KRA
```{r}
# clones no reps
LD_KRA <- popsub(LD_gd, "KRA")
ia(LD_KRA, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                
# 2.81044876 0.00100000 0.05679155 0.00100000 
```

```{r}
LD_KRA_MLG <- popsub(LD_gd_MLG, "KRA")
ia(LD_KRA_MLG, sample = 999)

#    Ia  p.Ia rbarD  p.rD 
#   NA    NA    NA    NA 
```

### NYN
```{r}
# clones no reps
LD_NYN <- popsub(LD_gd, "NYN")
ia(LD_NYN, sample = 999)

#        Ia      p.Ia     rbarD      p.rD                                                                    
# 35.498142  0.001000  0.189073  0.001000 
```

```{r}
LD_NYN_MLG <- popsub(LD_gd_MLG, "NYN")
ia(LD_NYN_MLG, sample = 999)

#        Ia       p.Ia      rbarD       p.rD                                                                                                                                           
# 9.53020134 0.00100000 0.06439325 0.00100000 
```

### BJO
```{r}
# clones no reps
LD_BJO <- popsub(LD_gd, "BJO")
ia(LD_BJO, sample = 999)

#         Ia       p.Ia      rbarD       p.rD                                                                                                                                           
# 0.91030882 0.00100000 0.01054078 0.00100000 
```

```{r}
LD_BJO_MLG <- popsub(LD_gd_MLG, "BJO")
ia(LD_BJO_MLG, sample = 999)

# Ia  p.Ia rbarD  p.rD 
#   NA    NA    NA    NA 
```
