---
title: "PCA"
output:
  html_notebook:
    toc: yes
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    toc_float: true
    theme: united  # many options for theme, this one is my favorites.
    highlight: tango  # specifies the syntax highlighting style
    code_folding: hide
#editor_options: 
  #chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Set up
```{r}
HOME <- "~/Library/CloudStorage/Dropbox/Steffi_Zostera_analysis_2023/Monitoring2023/Monitoring_eelgrass_scripts_for_figshare" # directory 
DATA <- paste0(HOME, "/data/") # data folder
RESULT <- paste0(HOME, "/results/") #result folder
SCRATCH <- paste0(HOME, "/scratch/") #result folder
```

```{r include=FALSE}
library(tidyverse)
library(adegenet)
library(vcfR)
library(fishualize)
library(patchwork)
```

# 25% missingnes

```{r}
load(paste0(SCRATCH, "gen_zostera_25_uniq.rda"))
head(gen_zostera_25_uniq)
gen_zostera_25_uniq@strata
```

First extract allelic data (as frequencies) and replace missing values using the accessor tab and then use the PCA procedure (dudi.pca).

```{r}
set.seed(9)
gen_zostera_25_scale <- scaleGen(gen_zostera_25_uniq, scale = F, NA.method = "mean")
gen_zostera_25_pca <- dudi.pca(gen_zostera_25_scale, cent = TRUE, scale = FALSE, scannf = FALSE, nf = 13)
```

## eigenvalues
A quick look at the eigenvalues

```{r}
barplot(gen_zostera_25_pca$eig)
```

```{r}
# calculate percentage from eigenvalues
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}

pc_percent  <- 1:10
for (i in 1:10) {
  pc_percent[i] <- paste0("PC",i," (", percent(
  gen_zostera_25_pca$eig[i]/sum(gen_zostera_25_pca$eig)), ")")
  }
pc_percent
```

```{r}
# setting colour
pop_col <- fish(n = 15, option = "Hypsypops_rubicundus")
# + scale_fill_fish_d(option = "Hypsypops_rubicundus")
  
#times -1 to flip the pca, 
gen_zostera_25_pca$li$Axis2 <- gen_zostera_25_pca$li$Axis2*-1
```

```{r}
# adding pop labels to the data
pca_zostera <-
  as_tibble(gen_zostera_25_pca$li) %>% 
  mutate(pop_code =pop(gen_zostera_25_uniq)) %>% 
  mutate(order =1:length(pop(gen_zostera_25_uniq)))
  #mutate(pop_name = factor(meta_df$site_full, levels = unique(meta_df$site_full)))
#  arrange(-order)

```

## PCA 1 & 2

```{r}
# plotting with ggglot
(pca1 <- ggplot(pca_zostera, aes(x=Axis1, y=Axis2, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.5) +
  labs(title ="Axis 1 & Axis 2", x=pc_percent[1], y=pc_percent[2]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  theme_light() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    #legend.justification=c(-0.2,-0.2), 
    legend.position="right"
  ))

ggsave(pca1, file = paste0(RESULT,"pca_fullname25.png"), width = 17, height = 15, units  ="cm", dpi=300)
```

```{r}
# plotting with ggglot
(pca <- ggplot(pca_zostera, aes(x=Axis1, y=Axis2, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.7) +
  labs(x=pc_percent[1], y=pc_percent[2]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  coord_cartesian(xlim = c(-8,5), ylim = c(-5, 8), expand = FALSE) +
  theme_light() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    legend.text = element_text(size = 11)
    #legend.justification=c(0,-0.55), 
    #legend.position=c(0.85,0.3)
  ))
#ggsave(pca, file = paste0(RESULT,"pca_fullname25.png"), width = 17, height = 15, units  ="cm", dpi=300)
```

```{r}
pca_1 <- ggplot(pca_zostera, aes(x=Axis1, fill=pop_code)) +
  geom_density(alpha=0.7, colour = NA) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  coord_cartesian(ylim = c(0,NA),xlim =c(-2.5, 11), expand = FALSE) +
  theme_void() +
  theme(
    axis.title = element_blank(),
    plot.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "none",
    axis.line.x = element_line(colour = "black")
  )
```

```{r}
pca_2 <- ggplot(pca_zostera, aes(x=Axis2, fill=pop_code)) +
  geom_density(alpha=0.7, colour = NA) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  coord_flip(ylim = c(0,NA), xlim =  c(-13, 8), expand = FALSE) +
  theme_void() +
  theme(
    axis.title = element_blank(),
    plot.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "none",
    axis.line.y = element_line(colour = "black")
  )
```

```{r, fig.height=7, fig.width=8}
layout <- "
AAAAA#
BBBBBC
BBBBBC
BBBBBC
BBBBBC
BBBBBC
"
pc_zostera_comb <- pca_1 + pca + pca_2 +
  plot_layout(design = layout, guides = "collect")
  #theme(legend.position='bottom')
pc_zostera_comb
ggsave(pc_zostera_comb, file = paste0(RESULT,"pca_zostera_comb25.png"), width = 18, height = 15, units  ="cm", dpi=300) 
```

## PCA Supplemetary
```{r , fig.height=7, fig.width=10}
# plotting with ggglot
pca1_2 <- ggplot(pca_zostera, aes(x=Axis1, y=Axis2, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.5) +
  labs(title ="Axis 1 & Axis 2", x=pc_percent[1], y=pc_percent[2]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  theme_light() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    legend.justification=c(0,0), 
    legend.position=c(0,0)
  )

eig <- ggplot(data.frame(gen_zostera_25_pca$eig),aes(seq_along(gen_zostera_25_pca$eig),gen_zostera_25_pca$eig)) +
  geom_bar(stat="identity") +
  labs(title = "Eigenvalues") +
  xlim(0,50) +
  theme_light() +
  theme(
  axis.title = element_blank(),
  plot.title = element_text(size = 20)
  )

library(patchwork)
(pca1_2 | eig) + plot_layout(guides = "collect") & theme(legend.text = element_text(size = 12))
ggsave(file = paste0(RESULT,"pca_supplementary25_2.0.png"), width = 18, height = 10, units  ="cm", dpi=300)
ggsave(file = paste0(RESULT,"pca_supplementary25_2.0.svg"), width = 18, height = 10, units  ="cm", dpi=300)
```

# Making plots with 81 MLLs
```{r}
load(paste0(SCRATCH, "gc_no_rep_mlg.rda"))
```

```{r}
# converting to genind
gen_zostera_mlg <- as(gc_no_rep_mlg, "genind")
```


First extract allelic data (as frequencies) and replace missing values using the accessor tab and then use the PCA procedure (dudi.pca).

```{r}
set.seed(9)
gen_zostera_25_scale <- scaleGen(gc_no_rep_mlg, scale = F, NA.method = "mean")
gen_zostera_25_pca <- dudi.pca(gen_zostera_25_scale, cent = TRUE, scale = FALSE, scannf = FALSE, nf = 13)
```

## eigenvalues
A quick look at the eigenvalues

```{r}
eig_81 <- barplot(gen_zostera_25_pca$eig)
ggsave(eig_81, file = paste0(RESULT,"eig_81.svg"), width = 10, height = 5, units  ="cm", dpi=300)
```

```{r}
# calculate percentage from eigenvalues
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}

pc_percent  <- 1:10
for (i in 1:10) {
  pc_percent[i] <- paste0("PC",i," (", percent(
    gen_zostera_25_pca$eig[i]/sum(gen_zostera_25_pca$eig)), ")")
}
pc_percent
```

```{r}
# setting colour
pop_col <- fish(n = 15, option = "Hypsypops_rubicundus")
# + scale_fill_fish_d(option = "Hypsypops_rubicundus")

#times -1 to flip the pca, 
gen_zostera_25_pca$li$Axis2 <- gen_zostera_25_pca$li$Axis2*-1
gen_zostera_25_pca$li$Axis1 <- gen_zostera_25_pca$li$Axis1*-1
```

```{r}
# adding pop labels to the data
pca_zostera <-
  as_tibble(gen_zostera_25_pca$li) %>% 
  mutate(pop_code =pop(gc_no_rep_mlg)) %>% 
  mutate(order =1:length(pop(gc_no_rep_mlg)))
#mutate(pop_name = factor(meta_df$site_full, levels = unique(meta_df$site_full)))
#  arrange(-order)
```

## PCA 1 & 2
```{r}
# plotting with ggglot
(pca1 <- ggplot(pca_zostera, aes(x=Axis1, y=Axis2, fill=pop_code, group = pop_code, shape = pop_code)) +
    geom_point(size=5, alpha=0.5) +
    labs(title ="Axis 1 & Axis 2", x=pc_percent[1], y=pc_percent[2]) +
    scale_fill_fish_d(option = "Hypsypops_rubicundus") +
    scale_color_fish_d(option = "Hypsypops_rubicundus") +
    scale_shape_manual(values= rep(21:25,4)) +
    stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
    theme_light() +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 15),
      plot.title = element_text(size = 20),
      legend.title = element_blank(),
      #legend.justification=c(-0.2,-0.2), 
      legend.position="right"
    ))
```


```{r}
# plotting with ggglot
(pca <- ggplot(pca_zostera, aes(x=Axis1, y=Axis2, fill=pop_code, group = pop_code, shape = pop_code)) +
    geom_point(size=4, alpha=0.7) +
    labs(x=pc_percent[1], y=pc_percent[2]) +
    scale_fill_fish_d(option = "Hypsypops_rubicundus") +
    scale_color_fish_d(option = "Hypsypops_rubicundus") +
    scale_shape_manual(values= rep(21:25,4)) +
    stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
    coord_cartesian(xlim = c(-7,4), ylim = c(-5.5, 5), expand = FALSE) +
    theme_light() +
    theme(
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 14),
      plot.title = element_text(size = 20),
    ))

ggsave(pca, file = paste0(RESULT,"pca_fullname25_MLG.png"), width = 17, height = 15, units  ="cm", dpi=300)
#ggsave(pca, file = paste0(RESULT,"pca_fullname25_MLG.svg"), width = 20, height = 15, units  ="cm", dpi=300)
```

```{r}
pca_1 <- ggplot(pca_zostera, aes(x=Axis1, fill=pop_code)) +
  geom_density(alpha=0.7, colour = NA) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  coord_cartesian(ylim = c(0,NA),xlim =c(-6,3.5), expand = FALSE) +
  theme_void() +
  theme(
    axis.title = element_blank(),
    plot.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "none",
    axis.line.x = element_line(colour = "black")
  )
```

```{r}
pca_2 <- ggplot(pca_zostera, aes(x=Axis2, fill=pop_code)) +
  geom_density(alpha=0.7, colour = NA) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  coord_flip(ylim = c(0,NA), xlim =  c(-3.5, 5), expand = FALSE) +
  theme_void() +
  theme(
    axis.title = element_blank(),
    plot.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "none",
    axis.line.y = element_line(colour = "black")
  )
```

```{r, fig.height=6, fig.width=7}
layout <- "
AAAAA#
BBBBBC
BBBBBC
BBBBBC
BBBBBC
BBBBBC
"
pc_zostera_comb <- pca_1 + pca + pca_2 +
  plot_layout(design = layout, guides = "collect")
#theme(legend.position='bottom')
pc_zostera_comb
ggsave(pc_zostera_comb, file = paste0(RESULT,"pca_zostera_comb25_MLG.png"), width = 18, height = 15, units  ="cm", dpi=300) 
```

## PCA 3-8

```{r , fig.height=7, fig.width=8}
# plotting with ggglot
pca1_3 <- ggplot(pca_zostera, aes(x=Axis1, y=Axis3, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.5) +
  labs(title ="Axis 1 & Axis 3", x=pc_percent[1], y=pc_percent[3]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  theme_light() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    legend.justification=c(0,0), 
    legend.position=c(0,0)
  )

# plotting with ggglot
pca3_4 <- ggplot(pca_zostera, aes(x=Axis3, y=Axis4*-1, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.5) +
  labs(title ="Axis 3 & Axis 4", x=pc_percent[3], y=pc_percent[4]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  theme_light() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    legend.justification=c(0,0), 
    legend.position=c(0,0)
  )

# plotting with ggglot
pca5_6 <- ggplot(pca_zostera, aes(x=Axis5, y=Axis6, fill=pop_code, group = pop_code, shape = pop_code)) +
  geom_point(size=5, alpha=0.5) +
  labs(title ="Axis 5 & Axis 6", x=pc_percent[5], y=pc_percent[6]) +
  scale_fill_fish_d(option = "Hypsypops_rubicundus") +
  scale_color_fish_d(option = "Hypsypops_rubicundus") +
  scale_shape_manual(values= rep(21:25,4)) +
  stat_ellipse(aes(col=pop_code), alpha=0.5, type = "norm", linetype = 2) +
  theme_light() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 20),
    legend.title = element_blank(),
    legend.justification=c(0,0), 
    legend.position=c(0,0)
  )

eig <- ggplot(data.frame(gen_zostera_25_pca$eig),aes(seq_along(gen_zostera_25_pca$eig),gen_zostera_25_pca$eig)) +
  geom_bar(stat="identity") +
  labs(title = "Eigenvalues") +
  xlim(0,50) +
  theme_light() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(size = 12)
  )

ggsave(eig, file = paste0(RESULT,"eig_supplementary25_MLG.svg"), width = 10, height = 5, units  ="cm", dpi=300)

library(patchwork)
(pca1_3 | pca3_4) / (pca5_6 | eig) + plot_layout(guides = "collect") & theme(legend.text = element_text(size = 20))
ggsave(file = paste0(RESULT,"pca_supplementary25_MLG.png"), width = 40, height = 40, units  ="cm", dpi=300)
```
